Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2015-11-28 tmm.tex
Change the colors in Fig.6 and Fig.7; also change some text

\section{Introduction}
\label{sec:intro}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
\IEEEPARstart{V}{ideo} streaming applications are becoming more and more popular on the Internet. Examples of the most common usage scenarios include online video-on-demand (VoD), live broadcasting, video chat, video-based online education, etc. In these applications, video data is constantly delivered to different clients over various transmission channels, and the client media player can begin playing the video before the entire file has been transmitted. Among other features, video streaming makes it easy to watch video anywhere and anytime, which represents the concept of Ubiquitous Multimedia Access (UMA), a very common topic within the multimedia community during the past few years \cite{Gualdi08}. Due to its convenience, video streaming also attracts a large number of users and becomes an increasingly important way for people to consume video content \cite{Chen13}.

In video streaming, the network condition significantly influences the user's watching experience. To provide better streaming service, some researchers choose to adopt the new software defined networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration of adaptive video manycast \cite{Xue15}. Although these studies offer a promising way to optimize network conditions, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the user to get best watching experience. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

@@

The advantage of simulcast systems like DASH is that they need little modification of existing infrastructure (servers, proxies, caches, firewall compatibility, etc.) and therefore are easy to deploy \cite{Bouten14}. Most commercial video websites, YouTube \cite{YouTube} for instance, use this option to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams significantly increase time and space cost. SVC streaming systems, on the contrary, do not have above shortcomings. Since an SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), it is able to provide seamless scalability and quality adaptation. Besides, performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast, despite some computation and bitrate overhead comparing with single layer stream. Therefore, to adopt SVC in the video streaming framework is a promising option and has attracted much attention in the study \cite{Chuah12}-\cite{Cicalo14}.

While these studies mostly focus on Quality of Service (QoS) optimization and resource allocation for specific networks, there are two key problems in the SVC adaptive video streaming system that need to be primarily addressed. First, as the prerequisite of the adaption to bandwidth change, the whole bitstream needs to be extracted so that sub-streams with various bitrates can be obtained. This is called the bitstream extraction problem. Optimal bitstream extraction algorithm aims to extract a sub-stream with best video quality, or minimal distortion, under the constraint of a given bitrate. Second, when the bandwidth changes with time, it's necessary to make decisions about whether and how to adjust the bitrate or video quality. This problem is called quality control, which means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. A good quality control algorithm should determine a suitable bitrate for current bandwidth condition, and keep the video quality both smooth and at a high level.

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the proposed system. The original DSS project is modified and SVC support is added to it, for the purpose of both research and application. Then novel algorithms for bitstream extraction and quality control are designed and implemented to achieve significant improvement in terms of user-perceived video quality. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective linear error model (LEM) is proposed and verified, which can be used to accurately estimate the distortion caused by discarding any combination of data packets from an SVC bitstream.
\item Second, on the basis of the distortion estimation enabled by LEM, a greedy-like algorithm is designed to assign priority for each data packet according to its R-D impact, thus enables optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) control concept is proposed to replace the native one in DSS, and achieves much lower quality fluctuation with higher quality level.
\end{itemize}

@@

A larger optimization window means each time we have more packets to compare and determine their priorities. Therefore, choosing a larger $K$ will result in more optimized extraction. However, doing this will also increase the computational complexity as well as the memory used to store the error vectors. Besides, if the 6-bit priority identifier field of the NALU header is used to store the priority values, only a range of 0$\sim$63 is supported, which also limits the value of $K$.

@@

\begin{table*}[t]
\centering
\caption{Algorithm Performance: Luma PSNR Gain Compared with JSVM}
\label{tab:results}
\begin{minipage}{0.95\linewidth}
% \renewcommand{\thefootnote}{\thempfootnote}
\centering
\begin{tabular}{c|c|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}| 
p{1.2cm}<{\centering}|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}}
%% |l|l| to left justify each column entry
%% |c|c| to center each column entry
%% use of \rule[]{}{} below opens up each row 
\hline \hline
\multicolumn{2}{c|}{Sequence} &
{\em Bus} & {\em City} & {\em Crew} & {\em Football} & {\em Foreman} & {\em Harbour} & {\em Mobile} & {\em Soccer} \\ \hline 
\multirow{2}{*}{JSVM QL\footnote{\label{footnote:JSVM_QL} Comparing the proposed algorithm with JSVM QL} (dB)}
& Max\footnote{\label{footnote:max} Maximum PSNR gain through all bitrate constraints} & 0.37 & 0.50 & 0.13 & 0.22 & 0.23 & 0.19 & 0.25 & 0.44 \\ \cline{2-10}
& Ave\footnote{\label{footnote:ave} Average PSNR gain through all bitrate constraints} & 0.11 & 0.11 & 0.05 & 0.12 & 0.05 & 0.05 & 0.06 & 0.13 \\ \hline
\multirow{2}{*}{JSVM no QL (dB)}
& Max & 0.50 & 0.83 & 0.32 & 0.29 & 0.50 & 0.34 & 0.61 & 0.53 \\ \cline{2-10}
& Ave & 0.18 & 0.37 & 0.20 & 0.15 & 0.22 & 0.15 & 0.25 & 0.29 \\ \hline
\end{tabular}
\end{minipage}
\end{table*}

@@

For simplicity and intuition, we combine the enhancement layers in the tested SVC video stream to construct quality levels (introduced in \ref{subsec:quality-level}) used in the quality control algorithm. The video stream is divided into several sub-streams, each of which is identified by the quality and temporal layer it contains, as illustrated in Table \ref{tab:sub-stream} (there are 3 layers in each scalability dimension; spatial scalability is not enabled). The quality level is defined as the sum of all sub-streams with ID equal or less than the level number. For example, quality level 3 would contain all sub-streams from 0 to 3, thus would include all temporal layers, but only the quality base layer 0. The tested stream has 9 quality levels from 0 to 8 (see Table \ref{tab:sub-stream}), and a maximum bit rate of 1587 kbps at level 8.

@@

\begin{figure*}[t]
\subfloat[Average quality level of different control algorithms]{
\includegraphics[width=0.5\textwidth]{Quality-all.png}
\label{fig:quality-all}}
\subfloat[Variance of quality level of different control algorithms]{
\includegraphics[width=0.5\textwidth]{Variance-all.png}
\label{fig:variance-all}}
\caption{Comparison of the long/short time prediction and the PDF algorithm in DSS}
\label{fig:performance-all}
\end{figure*}

\begin{figure*}[t]
\centering
\subfloat[Variance of quality level of PID and PDF control algorithms]{
\includegraphics[width=0.5\textwidth]{Quality-two.png}
\label{fig:quality-two}}
\subfloat[Average quality level of PID and PDF control algorithms]{
\includegraphics[width=0.5\textwidth]{Variance-two.png}
\label{fig:variance-two}}
\caption{Comparison of the proposed PID algorithm and the PDF algorithm in DDS}
\label{fig:performance-two}
\end{figure*}

\begin{table*}[t]
\centering
\caption{Improvement of Different Quality Control Schemes}
\label{tab:improvement}
\begin{tabular}[b]{p{4.2cm}<{\centering}|p{4.2cm}<{\centering}|p{4.2cm}<{\centering}}
\hline \hline
Quality control scheme & Average quality level & Quality variance \\ \hline
Packet delay feedback (PDF) & 0.0\% & 0.0\% \\ \hline
Short term prediction (STP) & \textbf{+13.5\%} & +38.5\% \\ \hline
Long term prediction (LTP) & +7.9\% & -17.2\% \\ \hline
PID based control scheme(PID) & +8.6\% & \textbf{-24.8\%} \\ \hline
\end{tabular}
\end{table*}

>>>>

\section{Introduction}
\label{sec:intro}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
\IEEEPARstart{V}{ideo} streaming applications are becoming more and more popular on the Internet. Examples of the most common usage scenarios include online video-on-demand (VoD), live broadcasting, video-based online education, etc. In these applications, video data is constantly delivered to different clients over various transmission channels, and the client media player can begin playing the video before the entire file has been transmitted. Among other features, video streaming makes it easy to watch video anywhere and anytime, which represents the concept of Ubiquitous Multimedia Access (UMA), a very common topic within the multimedia community during the past few years \cite{Gualdi08}. Due to its convenience, video streaming also attracts a large number of users and becomes an increasingly important way for people to consume video content \cite{Chen13}.

In video streaming, the network condition significantly influences the user's watching experience. To provide better streaming service, some researchers choose to adopt the new software defined networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration of adaptive video manycast \cite{Xue15}. Although these studies offer a promising way to optimize network conditions, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to cellular network). Even in the same network, the user's bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a constant bitrate. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

@@

Due to the easy deployment of simulcast systems \cite{Bouten14}, they are already used by many commercial video websites (e.g., YouTube \cite{YouTube}) to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams require a lot of computation time and disk space. In order to provide seamless quality adaptation at reasonable time and space cost, the adoption of SVC is necessary. An SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), and performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast. Therefore, to adopt SVC in the video streaming framework is technically a better option and has attracted much attention in the study \cite{Chuah12}-\cite{Cicalo14}.

There are two key problems in the SVC adaptive video streaming system that need to be primarily addressed. First, as the prerequisite of the adaption to bandwidth change, the whole bitstream needs to be extracted so that sub-streams with various bitrates can be obtained. This is called the bitstream extraction problem. Optimal bitstream extraction algorithm aims to extract a sub-stream with best video quality, or minimal distortion, under the constraint of a given bitrate. Second, when the bandwidth changes with time, it's necessary to make decisions about whether and how to adjust the bitrate or video quality. This problem is called quality control, which means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. A good quality control algorithm should determine a suitable bitrate for current bandwidth condition, and keep the video quality both smooth and at a high level.

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the proposed system. The original DSS project is modified and SVC support is added to it, for the purpose of both research and application. Then novel algorithms for bitstream extraction and quality control are designed and implemented to enable better quality of service for the user. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective linear error model (LEM) is proposed and verified, which can be used to accurately estimate the distortion caused by discarding any combination of data packets from an SVC bitstream.
\item Second, on the basis of the distortion estimation enabled by LEM, a greedy-like algorithm is designed to assign priority for each data packet according to its R-D impact, thus enables optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) control concept is proposed to replace the native one in DSS, and achieves much lower quality fluctuation with higher quality level.
\end{itemize}

@@

A larger optimization window means each time we have more packets to compare and determine their priorities. Therefore, choosing a larger $K$ has the potential of more optimized extraction. However, doing this will also increase the computational complexity as well as the memory used to store the error vectors. Besides, if the 6-bit priority identifier field of the NALU header is used to store the priority values, only a range of 0$\sim$63 is supported, which also limits the value of $K$.

@@

\begin{table*}[t]
\centering
\caption{Algorithm performance: Luma PSNR gain compared with JSVM}
\label{tab:results}
\begin{minipage}{0.95\linewidth}
% \renewcommand{\thefootnote}{\thempfootnote}
\centering
\begin{tabular}{c|c|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}| 
p{1.2cm}<{\centering}|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}|p{1.2cm}<{\centering}}
%% |l|l| to left justify each column entry
%% |c|c| to center each column entry
%% use of \rule[]{}{} below opens up each row
\hline \hline
\multicolumn{2}{c|}{Sequence} &
{\em Bus} & {\em City} & {\em Crew} & {\em Football} & {\em Foreman} & {\em Harbour} & {\em Mobile} & {\em Soccer} \\ \hline 
\multirow{2}{*}{JSVM QL\footnote{\label{footnote:JSVM_QL} Comparing the proposed algorithm with JSVM QL} (dB)}
& Max\footnote{\label{footnote:max} Maximum PSNR gain through all bitrate constraints} & 0.37 & 0.50 & 0.13 & 0.22 & 0.23 & 0.19 & 0.25 & 0.44 \\ \cline{2-10}
& Ave\footnote{\label{footnote:ave} Average PSNR gain through all bitrate constraints} & 0.11 & 0.11 & 0.05 & 0.12 & 0.05 & 0.05 & 0.06 & 0.13 \\ \hline
\multirow{2}{*}{JSVM no QL (dB)}
& Max & 0.50 & 0.83 & 0.32 & 0.29 & 0.50 & 0.34 & 0.61 & 0.53 \\ \cline{2-10}
& Ave & 0.18 & 0.37 & 0.20 & 0.15 & 0.22 & 0.15 & 0.25 & 0.29 \\ \hline
\end{tabular}
\end{minipage}
\end{table*}

@@

For simplicity and intuition, we combine the enhancement layers in the tested SVC video stream to construct quality levels (introduced in \ref{subsec:quality-level}) used in the quality control algorithm. The video stream is divided into several sub-streams, each of which is identified by the quality and temporal layer it contains, as illustrated in Table \ref{tab:sub-stream} (there are 3 layers in each scalability dimension; spatial scalability is not enabled). The quality level is defined as the sum of all sub-streams with ID equal or less than the level number. For example, quality level 3 would contain all sub-streams from 0 to 3, thus would include all temporal layers, but only the quality base layer 0. The tested stream has 9 quality levels from 0 to 8 (see Table \ref{tab:sub-stream}), and a maximum bitrate of 1587 kbps at level 8.

@@

\begin{figure*}[t]
\subfloat[Average quality level of different control algorithms]{
\includegraphics[width=0.5\textwidth]{Quality-all.png}
\label{fig:quality-all}}
\subfloat[Quality level variance of different control algorithms]{
\includegraphics[width=0.5\textwidth]{Variance-all.png}
\label{fig:variance-all}}
\caption{Comparison of the long/short time prediction and the PDF algorithm in DSS}
\label{fig:performance-all}
\end{figure*}

\begin{figure*}[t]
\centering
\subfloat[Average quality level of PID and PDF control algorithms]{
\includegraphics[width=0.5\textwidth]{Quality-two.png}
\label{fig:quality-two}}
\subfloat[Quality level variance of PID and PDF control algorithms]{
\includegraphics[width=0.5\textwidth]{Variance-two.png}
\label{fig:variance-two}}
\caption{Comparison of the proposed PID algorithm and the PDF algorithm in DDS}
\label{fig:performance-two}
\end{figure*}

\begin{table*}[t]
\centering
\caption{Improvement of different quality control algorithms}
\label{tab:improvement}
\begin{tabular}[b]{p{4.2cm}<{\centering}|p{4.2cm}<{\centering}|p{4.2cm}<{\centering}}
\hline \hline
Quality control algorithm & Average quality level & Quality variance \\ \hline
Packet delay feedback (PDF) & 0.0\% & 0.0\% \\ \hline
Short term prediction (STP) & \textbf{+13.5\%} & +38.5\% \\ \hline
Long term prediction (LTP) & +7.9\% & -17.2\% \\ \hline
PID based control scheme(PID) & +8.6\% & \textbf{-24.8\%} \\ \hline
\end{tabular}
\end{table*}
